Topological Equivariant Artist Model for Visualization Library Architecture

### 2023-05-08 paper/main.tex
intro + rw

\usepackage{notation} %notation conventions
% correct bad hyphenation here
\hyphenation{}

@@

We use category theory and algebraic structures because they can be directly translated into software design specifications and our goal is to improve visulization tools. The algebraic structures can be translated into analogous programmatic types, while category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

@@

 \begin{definition}
$f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant, pittsNominalSetsNames2013}.
 \end{definition}

 \note{add something back in for homomorphic}

@@

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
 Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings. \note{add in Scales/APT/ GoG/Hullman/ETC}

@@

\section{Internel Library Specification}

>>>>

\usepackage{notation} %notation conventions
% correct bad hyphenation here

@@

We use category theory and algebraic structures because they can be directly translated into software design specifications and our goal is to improve visualization tools. The algebraic structures can be translated into analogous programmatic types, while category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

@@

 \begin{definition}
$f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant, pittsNominalSetsNames2013}.
 \end{definition}
In an equivariant transform, mapping the data after acting on it $f(act)$ is equivalent to acting on the data after mapping it $act(f)$ since $G$ acts on both. A homomorphism does not require that the input and output are of the same type; instead

 \begin{definition}
\note{must sort out exact appropriate definition of group homomorphism here}
 \end{definition}

@@

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
 Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings. Steven's measurement scale's and the structure preservation they define is the basis of Wilkenson's scales\cite{wilkinsonGrammarGraphics2005} in Grammar of Graphics and derivative libraries, and in this work we generalize the concept of fields to support a wider array of field types. This also allows us to specify what consistency means when creating different views of data from a single dataset, as discussed in Qu and Hullman \cite{hullmanKeeping2018}, specifically in terms of each field type separate from the overall structure.

@@

\pagebreak
\section{Internel Library Specification}
